---
title: "Practical Machine Learning Project"
author: "Gracy"
date: "22 December 2017"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache=TRUE)
```


###OVERVIEW

This to predict the manner in which 6 participants performed some exercise as described below. This is the "classe" variable in the training set. The machine learning algorithm described here is applied to the 20 test cases available in the test data and the predictions are submitted in appropriate format to the Course Project Prediction Quiz for automated grading.
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).



###Environment Setting and Read files into R
```{r freememory}
rm(list=ls()) 
```

```{r wd}
#Set the working environment. Switch to required directory
setwd("C:/Users/Gracy/Coursera - Data Science Specialization/Course 8 - Practical Machine Learning/Week 4")
```

```{r url}
fileUrl.train <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
fileUrl.test <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

```{r downloadfile}
#download.file(fileUrl.train,destfile = "pml-training.csv")
#download.file(fileUrl.test,destfile = "pml-testing.csv")
```


```{r load_lib,warning=FALSE}
library(caret)
library(parallel)
library(doParallel)
library(dplyr)
```

```{r readfiles}
training <- read.csv("pml-training.csv")
testing <- read.csv("pml-testing.csv")
dim(testing)
dim(training)
```

###Clean up the Data

```{r remove_NA}
# remove variables with Nearly Zero Variance
NZV <- nearZeroVar(training)
training <- training[, -NZV]
testing  <- testing[, -NZV]
dim(training)
dim(testing)
```

```{r removeNA}
# remove variables that are mostly NA
AllNA    <- sapply(training, function(x) mean(is.na(x))) > 0.95
training <- training[, AllNA==FALSE]
testing  <- testing[, AllNA==FALSE]
dim(training)
dim(testing)
```

```{r removeidentification}
# remove identification only variables (columns 1 to 5)
training <- training[, -(1:5)]
testing  <- testing[, -(1:5)]
dim(training)
dim(testing)
```

### Making an initial fit with all predictors for feature selection
To improve processing time of the multiple executions of the train() function, caret supports the parallel processing capabilities of the parallel package. Parallel processing in caret can be accomplished with the parallel and doParallel packages. 
```{r random_forest_fit}
#Method: Random Forest
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)

fitControl <- trainControl(method = "cv", number = 5, allowParallel = TRUE)

fit <- train(classe ~ ., method="rf",data=training,trControl = fitControl)

stopCluster(cluster)
registerDoSEQ()
```

###Feature Selection using Variable Importance method

The varImp process chooses the best 20 predictors and those are the columns selected for testing and training data.
```{r variable_importance}
var.imp <- varImp(fit)
var.imp
var.imp.cols.train <- c("num_window","roll_belt", "pitch_forearm","yaw_belt",     "magnet_dumbbell_z","pitch_belt",      "magnet_dumbbell_y",
"roll_forearm",     "accel_dumbbell_y", "magnet_dumbbell_x","accel_forearm_x",  "roll_dumbbell",    
"accel_belt_z",     "total_accel_dumbbell","accel_dumbbell_z", "magnet_belt_z",          "magnet_forearm_z",       
"magnet_belt_y",          "magnet_belt_x",          "roll_arm" , "classe")
var.imp.cols.test <- c("num_window","roll_belt", "pitch_forearm","yaw_belt",     "magnet_dumbbell_z","pitch_belt",      "magnet_dumbbell_y",
"roll_forearm",     "accel_dumbbell_y", "magnet_dumbbell_x","accel_forearm_x",  "roll_dumbbell",    
"accel_belt_z",     "total_accel_dumbbell","accel_dumbbell_z", "magnet_belt_z",          "magnet_forearm_z",       
"magnet_belt_y",          "magnet_belt_x",          "roll_arm" )
training <- training[,c(var.imp.cols.train)]
testing <- testing[,c(var.imp.cols.test)]

plot(var.imp)
```


###Best Fit
This best fit method of leaps library calculates the best predictors using residual sum of squares measure. We find that after the first 15 features the remaining variables are not of much importance. So we proced on to the final fit methods with these 15 predictors.

```{r best_fit}
#best fit
library(leaps)
# Perform a best fit
bestFit=regsubsets(classe~.,training,nvmax=15)

# Generate a summary of the fit
bfSummary=summary(bestFit)

# Plot the Residual Sum of Squares vs number of variables 
plot(bfSummary$rss,xlab="Number of Variables",ylab="RSS",type="l",main="Best fit RSS vs No of features")
# Get the index of the minimum value

a=which.min(bfSummary$rss)
# Mark this in red
points(a,bfSummary$rss[a],col="red",cex=2,pch=20)
#The plot below shows that the Best fit occurs with all 15 features included. Notice that there is no significant change in RSS from 15 features onward.
```


###Random Forest ( with final set of predictors)
```{r random_forest_fit_final}
#Method: Random Forest
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)

fitControl <- trainControl(method = "cv", number = 5, allowParallel = TRUE)

finalfit1 <- train(classe ~ ., method="rf",data=training,trControl = fitControl)

stopCluster(cluster)
registerDoSEQ()

result1 <- confusionMatrix(finalfit1)
result1
plot(result1$table, col = result1$byClass, main = "Random Forest")
```


###Decision Trees
```{r decision_trees}
#Method: Decision Trees
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)

fitControl <- trainControl(method = "cv", number = 5, allowParallel = TRUE)

finalfit2 <- train(classe ~ ., method="rpart",data=training,trControl = fitControl)

stopCluster(cluster)
registerDoSEQ()
result2 <- confusionMatrix(finalfit2)
result2
plot(result2$table, col = result2$byClass, main = "Decision Trees")
```



###Gradient Boosting
```{r gradient_boosting}
#Method: Gradient Boost
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)

fitControl <- trainControl(method = "cv", number = 5, allowParallel = TRUE)

finalfit3 <- train(classe ~ ., method="gbm",data=training,trControl = fitControl)

stopCluster(cluster)
registerDoSEQ()
result3 <- confusionMatrix(finalfit3)
result3
plot(result3$table, col = result3$byClass, main = "Gradient Boost")
```

The accuracy of the above 3 regression modeling methods are:

Random Forest : 0.9979
Decision Tree : 0.544
GBM : 0.9924

####PREDICTION with best model - Random Forest
The required accuracy for this project is achieved from Random Forest. Therefore predicting with that model.
```{r predict}
prediction <- predict(finalfit1,newdata = testing)
prediction

```

####Out of Sample Error


```{r sample_error}
outOfSampleError.accuracy <- sum(prediction == testing$classe)/length(prediction)

outOfSampleError.accuracy

outOfSampleError <- 1 - outOfSampleError.accuracy

outOfSampleError
```

